---
title: "Assignment 3 - Causal inference"
author: "Astrid Rybner, Kata Molnar, Sofie Rødkjær and Nicole Dwenger"
date: "2/5/2020"
output:   
  md_document:
    variant: markdown_github
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 3 - Exploring causal inference issues

In this assignment we explore some issues related to multiple regressions (regressions with more than one predictor), and inferred (causal) relations between variables. N.B. the data is simulated (to make sure I know the actual mechanism generating it), but it's based on a real study. So bear with a longish introduction to get into the details of what we are doing and why it is important.

### Altercentric intrusion in schizophrenia

People with schizophrenia often report altered control and distinction of self-other representations: intrusive thoughts, hearing of voices, delusions of mind reading, paranoia, etc (a substantial portion of the psychotic symptoms experienced in schizophrenia). These have been variously attributed to hypermentalizing (over attribution of mental states to others), social impairment (over preoccupation with own thought processes), hyper socialization (inability to inhibit information from others), etc.

The current study investigates 1) whether schizophrenia is indeed related to altered control and distinction of self-other representations, in particular altercentric intrusions (inability to inhibit social information), and 2) whether these are related to the relevant psychotic symptoms. N.B. the actual study also investigates egocentric intrusion, do check the papers below if interested.

The task is a slightly modified version of this: https://www.ncbi.nlm.nih.gov/pubmed/20731512 You look at a picture with some dots visible to you, as well as with a different person with a different set of dots visible to them. The number of dots you see and that the other sees can be the same (congruent condition) or not (incongruent condition). You are tasked to indicate whether a given number (e.g. 3) matches the number of dots you see (and the dots visible to the other person are irrelevant to the task).


The tasks investigates altercentric intrusion: will your reaction time change according to whether the other person is seeing the same amount of dots as you, or not? The idea is that if you correctly inhibit social information, your reaction time should not change, as the information about the other person is not relevant. On the contrary, if you nevertheless use task irrelevant social information, you'll be slower at indicating whether 3 is the right number of dots when the other person sees a different amount of dots than you (conflicting information).
The bigger the difference between RTs in the congruent and incongruent condition the bigger the altercentric intrusion effect.

For each participant you have 6 variables: 1) ID, 2) AltercentricIntrusion (continuous score), 3) Diagnosis (schizophrenia vs. control), 4) VoiceHearing (severity of voice hearing symptoms, continuous score of the severity of the symptom as measured by a clinician), 5) MindReading (severity of delusions of mind reading, continuous score of the severity of the symptom as measured by a clinician); 6) Apathy (severity of lack of motivation in taking care of oneself, from washing to showing up at work, continuous score of the severity of the symptom as measured by a clinician).

The research questions you have to answer are the following:

## First part

Q1.1) Does schizophrenia involved altercentric intrusion? Define model and priors. Test the implications of your priors (prior predictive checks) and if needed adjust them. Run the model. Test the quality of the fitted model (posterior predictive checks). Assess the evidence in favor of an increased altercentric intrusion in schizophrenia. Report the model and the results, including plots.

```{r}
#1:test h that scz has higher ACI than controls 

pacman::p_load(tidyverse, brms, dplyr)
pacman::p_load(patchwork)

#PREPARE DATA -------------------------------------

#load data
d <- read_csv("Ass3.csv")

#check if continuous 
summary(d)

#tell data set what 0 and 1 mean 
d$Diagnosis <- plyr::revalue(as.character(d$Diagnosis), 
                             c("0"="Controls", "1"="Schizophrenia"))

#add factors
d <- d %>%
  mutate(
    ID = as.factor(ID),
    Diagnosis = as.factor(Diagnosis)
  )


#DEFINE FORMULA -------------------------------------

#no repeated measures,bf = bayesian formula
AltercentricDiagnosis_f0 <- bf(
  AltercentricIntrusion ~ 1 + Diagnosis
)

#give absolute values for both controls and SCZ
AltercentricDiagnosis_f <- bf(
  AltercentricIntrusion ~ 0 + Diagnosis
)

#PRIORS -------------------------------------

#find out what we can define
get_prior(AltercentricDiagnosis_f0, family = gaussian, d)
get_prior(AltercentricDiagnosis_f, family = gaussian, d)

#define priors
priorDiagnosis <- c(
  prior(normal(4, 1), class = b), # if only b long tail - to avoid: sigma 
  prior(normal(1, 2), class = sigma)
) 

#test priors
AltercentricDiagnosis_PriorCheck_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only"
)

#prior predictive checks
pp_check(AltercentricDiagnosis_PriorCheck_m, nsamples = 100)


#RUN MODEL -------------------------------------
AltercentricDiagnosis_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T
)

#posterior predictive check
pp_check(AltercentricDiagnosis_m, nsamples = 100)


#MODEL OUTPUT + PLOTS -------------------------------------

#model output, check warnings
AltercentricDiagnosis_m
summary(AltercentricDiagnosis_m)

hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > 0")

hypothesis(AltercentricDiagnosis_m,
           "DiagnosisControls > 0")

#hypotheis testing, updating check
hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > DiagnosisControls")
plot(hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > DiagnosisControls"))

hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > DiagnosisControls")
#evidence ratio: how many of the samples are compatible with our hypotheiss, out of 4000 samples (here) - over evidence 

#plot effect
conditional_effects(AltercentricDiagnosis_m)
plot(conditional_effects(AltercentricDiagnosis_m), points=T)

```

The model indicates a credible difference in altercentric intrusion in the two groups supporting our hypothesis (b = 0.36, CIs = 0.16, 0.57, ER = 1332). Controls showed on average an altercentric intrusion effect of 3.86 (CIs 3.74, 3.98), and schizophrenia of 4.22 (CIs = 4.01, 4.43).
[Add plot of the effects]

# SI
The model had no divergences, a Rhat of 1, and Effective Sample Sizes above 2000 for both Bulk and Tail.
[Add prior and posterior checks plots; add updating check plot]

Q1.2) Is altercentric intrusion related to specific symptoms *in the patients*? Identify which of the symptoms could be relevant. Should you include more than one symptom? Build models, priors, predictive checks. Assess the evidence and report models and results, including plots. Discuss whether the results make sense.
```{r}
#if there is a true difference in AIC then it might be true that some of the symptoms might be related to the confusion of representation of self and other

#PREPARE DATA -------------------------------------

#scale - if we don't there will be some correlations of intercept and slope 
d <- d %>%
  mutate(
    AltercentricIntrusion_truescale = scale(AltercentricIntrusion),
    VoiceHearing = scale(VoiceHearing),
    MindReading = scale(MindReading),
    Apathy = scale(Apathy)
  )

d <- d %>%
  mutate(
    AltercentricIntrusion_scale = (AltercentricIntrusion - mean(AltercentricIntrusion)) / sd (AltercentricIntrusion))

#subset with only schizophrenia 
scz <- subset(d, Diagnosis == "Schizophrenia")
summary(scz)
sd(scz$AltercentricIntrusion)

#DEFINE FORMULA -------------------------------------
#voice hearing
voicehearing_f <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing
)

#mind reading
mindreading_f <- bf(
  AltercentricIntrusion ~ 1 + MindReading
)

#apathy
apathy_f <- bf(
  AltercentricIntrusion ~ 1 + Apathy
)

#voice + mind 
voicemind_f <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing + MindReading
)

#voice + mind + apathy
all_f <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing + MindReading + Apathy
)

#PRIORS -------------------------------------
#get priors
get_prior(voicehearing_f, family = gaussian, scz)
get_prior(mindreading_f, family = gaussian, scz)
get_prior(apathy_f, family = gaussian, scz)
get_prior(voicemind_f, family = gaussian, scz)
get_prior(all_f, family = gaussian, scz)

#define priors
prior <- c(
  prior(normal(0,.3), class = b), #make 1 smaller? 
  prior(normal(1,2), class = sigma),
  prior(normal(0,1), class = Intercept)) #0 becasue we scaled things

#test prios
voicehearing_PriorCheck_m <- brm(
  formula = voicehearing_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = "only"
)

#prior checks
mindreading_PriorCheck_m <- brm(
  formula = mindreading_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = "only"
)

apathy_PriorCheck_m <- brm(
  formula = apathy_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = "only"
)

voicemind_PriorCheck_m <- brm(
  formula = voicemind_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = "only"
)

all_PriorCheck_m <- brm(
  formula = all_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = "only"
)

#prior predictive checks
pp_check(voicehearing_PriorCheck_m, nsamples = 100)
pp_check(mindreading_PriorCheck_m, nsamples = 100)
pp_check(apathy_PriorCheck_m, nsamples = 100)
pp_check(voicemind_PriorCheck_m, nsamples = 100)

#RUN MODELS -------------------------------------
voicehearing_m <- brm(
  formula = voicehearing_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "voicehearing_m"
)

mindreading_m <- brm(
  formula = mindreading_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "mindreading_m"
)

apathy_m <- brm(
  formula = apathy_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "apathy_m"
)

voicemind_m <- brm(
  formula = voicemind_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "voicemind_m"
)

all_m <- brm(
  formula = all_f,
  data = scz,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "all_model"
)

#posterior predictive checks
pp_check(voicehearing_m, nsamples = 100)
pp_check(mindreading_m, nsamples = 100)
pp_check(apathy_m, nsamples = 100)
pp_check(voicemind_m, nsamples = 100)

#MODEL OUTPUT AND PLOTS -------------------------------------
#model output
summary(voicehearing_m)
summary(mindreading_m)
summary(apathy_m)
summary(voicemind_m)
summary(all_m)

#hypothesis 
hypothesis(voicehearing_m, "VoiceHearing > 0")
hypothesis(mindreading_m, "MindReading > 0")
hypothesis(apathy_m, "Apathy = 0")
hypothesis(apathy_m, "Apathy < 0")
hypothesis(voicemind_m, c("VoiceHearing > 0", "MindReading > 0"))
hypothesis(all_m, c("VoiceHearing > 0", "MindReading > 0", "Apathy = 0"))
hypothesis(all_m, c("VoiceHearing > 0", "MindReading > 0", "Apathy < 0"))

### PLOTS
#overall
plot(voicehearing_m)
plot(mindreading_m)
plot(apathy_m)
plot(voicemind_m)
plot(all_m)

#hypothesis
plot(hypothesis(voicehearing_m,"VoiceHearing > 0"))
plot(hypothesis(mindreading_m,"MindReading > 0"))
plot(hypothesis(apathy_m,"Apathy = 0"))
plot(hypothesis(voicemind_m, "VoiceHearing + MindReading > 0"))
plot(hypothesis(voicemind_m, c("VoiceHearing > 0", "MindReading > 0")))
hypothesis(all_m, c("VoiceHearing > 0", "MindReading > 0", "Apathy < 0"))

#Conditional Effects
conditional_effects(voicehearing_m)
conditional_effects(mindreading_m)
conditional_effects(apathy_m)
conditional_effects(voicemind_m)
conditional_effects(all_m)

plot(conditional_effects(voicehearing_m), points=T)
plot(conditional_effects(mindreading_m), points=T)
plot(conditional_effects(apathy_m), points=T)
plot(conditional_effects(voicemind_m), points=T)
plot(conditional_effects(all_m), points=T)

#annoying: goes against hypothesis 
```

```{r}
#MODEL COMPARISON
#maybe we just overfitted - find the model that gives the best result and is not overfitting
voicehearing_m <- add_criterion(voicehearing_m, criterion = "loo")
mindreading_m <- add_criterion(mindreading_m, criterion = "loo")
apathy_m <- add_criterion(apathy_m, criterion = "loo")
voicemind_m <- add_criterion(voicemind_m, criterion = "loo")
all_m <- add_criterion(all_m, criterion = "loo")

loo_compare(voicehearing_m,
            mindreading_m,
            apathy_m,
            voicemind_m,
            all_m)

loo_model_weights(voicehearing_m,
            mindreading_m,
            apathy_m,
            voicemind_m,
            all_m)

```


## Second part

Q2.1) However, we know that the diagnosis is based on symptom assessment: if the overall sum of symptoms is severe enough, the participant gets a diagnosis. In other words, by selecting the patients, and including the symptoms in the model we might have inadvertently introduced an issue in our inference. Do try to draw a causal graph (Directed Acyclical Graph) of the variables and compare it with the types of causal graphs presented in the slides. Discuss which biases you might have introduced.

Q2.2.) Redesign your analysis following the graph and report how the results change

```{r}
#we know that diagnosis is based on symptom assessment 
#by selecting the patients and including the symptoms we might have introduced an issue 

#formulas are still the same

#PRIORS -------------------------------------
#get priors
get_prior(voicehearing_f, family = gaussian, d)
get_prior(mindreading_f, family = gaussian, d)
get_prior(apathy_f, family = gaussian, d)
get_prior(voicemind_f, family = gaussian, d)
get_prior(all_f, family = gaussian, d)

#define prior
prior <- c(
  prior(normal(0,.3), class = b), #make 1 smaller? 
  prior(normal(1,2), class = sigma),
  prior(normal(0,1), class = Intercept)) #0 becasue we scaled things

#test priors
voicehearing_PriorCheck_m1 <- brm(
  formula = voicehearing_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = "only",
  file = "voicehearing_PriorCheck_m1"
)

mindreading_PriorCheck_m1 <- brm(
  formula = mindreading_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = "only",
  file = "mindreading_PriorCheck_m1"
)

apathy_PriorCheck_m1 <- brm(
  formula = apathy_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = "only",
  file = "apathy_PriorCheck_m1"
)

voicemind_PriorCheck_m1 <- brm(
  formula = voicemind_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = "only",
  file = "voicemind_PriorCheck_m1"
)

all_PriorCheck_m1 <- brm(
  formula = all_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = "only", 
  file = "all_PriorCheck_m1"
)

#prior predictive checks
pp_check(voicehearing_PriorCheck_m1, nsamples = 100)
pp_check(mindreading_PriorCheck_m1, nsamples = 100)
pp_check(apathy_PriorCheck_m1, nsamples = 100)
pp_check(voicemind_PriorCheck_m1, nsamples = 100)


#RUN MODEL -------------------------------------
voicehearing_m1 <- brm(
  formula = voicehearing_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "voicehearing_m1"
)

mindreading_m1 <- brm(
  formula = mindreading_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "mindreading_m1"
)

apathy_m1 <- brm(
  formula = apathy_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "apathy_m1"
)

voicemind_m1 <- brm(
  formula = voicemind_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "voicemind_m1"
)

all_m1 <- brm(
  formula = all_f,
  data = d,
  family = gaussian,
  prior = prior,
  sample_prior = T, 
  file = "all_model1"
)

#posterior predictive checks
pp_check(voicehearing_m1, nsamples = 100)
pp_check(mindreading_m1, nsamples = 100)
pp_check(apathy_m1, nsamples = 100)
pp_check(voicemind_m1, nsamples = 100)


#RUN MODELS AND PLOTS -------------------------------------
summary(voicehearing_m1)
summary(mindreading_m1)
summary(apathy_m1)
summary(voicemind_m1)
summary(all_m1)

#hypothesis 
hypothesis(voicehearing_m1, "VoiceHearing > 0")
hypothesis(mindreading_m1, "MindReading > 0")
hypothesis(apathy_m1, "Apathy = 0")
hypothesis(apathy_m1, "Apathy < 0")
hypothesis(voicemind_m1, c("VoiceHearing > 0", "MindReading > 0"))
hypothesis(all_m1, c("VoiceHearing > 0", "MindReading > 0", "Apathy = 0"))
hypothesis(all_m1, c("VoiceHearing > 0", "MindReading > 0", "Apathy < 0"))

### PLOTS
#overall
plot(voicehearing_m1)
plot(mindreading_m1)
plot(apathy_m1)
plot(voicemind_m1)
plot(all_m1)

#hypothesis
plot(hypothesis(voicehearing_m1,"VoiceHearing > 0"))
plot(hypothesis(mindreading_m1,"MindReading > 0"))
plot(hypothesis(apathy_m1,"Apathy = 0"))
plot(hypothesis(voicemind_m1, c("VoiceHearing > 0", "MindReading > 0")))
hypothesis(all_m1, c("VoiceHearing > 0", "MindReading > 0", "Apathy < 0"))

#Conditional Effects
conditional_effects(voicehearing_m1)
conditional_effects(mindreading_m1)
conditional_effects(apathy_m1)
conditional_effects(voicemind_m1)
conditional_effects(all_m1)

plot(conditional_effects(voicehearing_m1), points=T)
plot(conditional_effects(mindreading_m1), points=T)
plot(conditional_effects(apathy_m1), points=T)
plot(conditional_effects(voicemind_m1), points=T)
plot(conditional_effects(all_m1), points=T)
```

```{r}
#MODEL COMPARISON
voicehearing_m1 <- add_criterion(voicehearing_m1, criterion = "loo")
mindreading_m1 <- add_criterion(mindreading_m1, criterion = "loo")
apathy_m1 <- add_criterion(apathy_m1, criterion = "loo")
voicemind_m1 <- add_criterion(voicemind_m1, criterion = "loo")
all_m1 <- add_criterion(all_m1, criterion = "loo")

loo_compare(voicehearing_m1,
            mindreading_m1,
            apathy_m1,
            voicemind_m1,
            all_m1)

loo_model_weights(voicehearing_m1,
            mindreading_m1,
            apathy_m1,
            voicemind_m1,
            all_m1)
```

## Third part

These issues are very difficult to think through, and not knowing the causal mechanisms generating the data in advance makes our inferences even more unreliable. To explore these issues, I recommend using simulations. In other words, defining a "true" model, generating data from it and assessing what different analyses would lead you to infer (and therefore which biases they might introduce). You can find the code I used to simulate your data below.

Q3.1) Look through the code and identify whether the results you have match the underlying truth. Discuss what you have learned.

Q3.2) OPTIONAL: is this a general pattern? Try varying the parameters (e.g. correlation values) and assess whether the new dataset(s) leads to the same biases in your analysis.


```{r}
#most of the people will first try to look at the differences 
#and then look at symptoms only within patients - which might be confound
#might want to see things more realistically 
#maybe also simulate that the scale is bad for low symptoms 
#add noise in different levels of symptoms 
#look at network of correlation between symptoms 
#should question regression analysis symptoms + cognitive effects + thinking that model comparison won't leave you anywhere if you don't have a theory 

pacman::p_load(MASS, tidyverse, psych)

seed <- 1981 # Defining a seed so the results are always the same
n <- 300 # Defining the amount of participants

SymptomCorr <- .2 # Defining the correlation of symptoms (as they tend to co-occur)
EffectCorrRel <- .2 # Defining the correlation between relevant symptoms and effect (Some symptoms are positively correlated with the effect)
EffectCorrIrrel <- 0 # Defining the correlation between irrelevant symptoms and effect (none)

# Creating the variance-covariance matrix for the variables we want to generate (3 symptoms, 1 effect)
Sigma <- matrix(data=c(1,SymptomCorr,SymptomCorr,EffectCorrRel,
                       SymptomCorr,1,SymptomCorr,EffectCorrRel,
                       SymptomCorr,SymptomCorr,1,EffectCorrIrrel,
                       EffectCorrRel,EffectCorrRel,EffectCorrIrrel,1),
                       nrow=4,ncol=4)

Sigma

## Generate data from a multivariate (mvr) normal (n) distribution
d <- mvrnorm(n = n, # number of participant
        mu = c(1.2, 1.2, 1.2, 4), # mean of each variable
        Sigma) # variance co-variance matrix

# Giving meaningful names to variables and add ID
d <- data.frame(
  VoiceHearing = d[,1], 
  MindReading =  d[,2],
  Apathy =  d[,3], 
  AltercentricIntrusion = d[,4],
  ID = seq(nrow(d)))

# Assessing whether the participant has schizophrenia (high enough sum of symptoms)
# Here we choose participants scoring above 75% percentile (the most severe ones)
d$Diagnosis <- 0
d$Diagnosis[(d$VoiceHearing + d$MindReading + d$Apathy) > 
              quantile(d$VoiceHearing + d$MindReading + d$Apathy, .75)] <-1

## Plotting the relation between variables in schizophrenia
d1 <- d %>% subset(Diagnosis==1) %>% dplyr::select(-Diagnosis, -ID)
pairs.panels(d1)

## Plotting the relation between variables all participants
pairs.panels(dplyr::select(d,-Diagnosis, -ID))


#write_csv(d, "data/Ass3.csv")
```

